---
title: "STAT40830-assignment1"
author: "Ye Ji"
format:
  revealjs:
    slide-number: true
    theme: simple
    toc: true
    css: individuation.css
editor: visual
---

## Introduction

This analysis presents the economic, social and environmental indicator data of three countries—Canada, New Zealand, Cuba.The objective is to examine how these nations differ in key development metrics over time.

<!-- Added on 2025-6-17 -->

This report explores cross-country development indicators.

## Motivation

Chose Canada, New Zealand, and Cuba to highlight differences in regional development trajectories. Canada and New Zealand are high-income economies with stable growth, while Cuba offers a unique case.

The goal is to compare social and economic indicators in these diverse contexts.

## Data Overview

-   Total records: \~210,000.
-   Time range: 1960–2023.
-   Over 100 unique indicators per country.
-   Format: country, year, indicator, value.

## Code

```{r}
library(data.table)

# Read CSVs and skip the first metadata row
nzl <- fread("indicators_nzl.csv", skip = 1)
can <- fread("indicators_can.csv", skip = 1)
cub <- fread("indicators_cub.csv", skip = 1)

# Add country label for grouping
nzl[, Country := "New Zealand"]
can[, Country := "Canada"]
cub[, Country := "Cuba"]
```

### **Preprocess**

```{r}
# Merge into one table
all_data <- rbindlist(list(nzl, can, cub))

# Rename columns for clarity
setnames(all_data, old = names(all_data), new = c("CountryName", "ISO3", "Year", "Indicator", "Code", "Value", "Country"))

# Convert to correct types
all_data[, Year := as.integer(Year)]
all_data[, Value := as.numeric(Value)]

```

```{r}
# List unique indicators
head(unique(all_data$Indicator), 10)
```

**We select two key indicators(GDP per capita, Life expectancy at birth) for further analysis.**

## GDP per Capita

The following table shows the average GDP per capita for each country from 1960 to 2023.\
This helps us identify long-term economic trends and differences in wealth levels across the selected countries.

```{r}
gdp <- all_data[Indicator == "GDP per capita (current US$)"]

# Mean GDP per country and year
gdp_summary <- gdp[, .(MeanGDP = mean(Value, na.rm = TRUE)), keyby = .(Country, Year)]
gdp_summary

```

## Plot (GDP per Capita):

This plot displays how GDP per capita has changed over time.

```{r}
library(ggplot2)

ggplot(gdp, aes(x = Year, y = Value, color = Country)) +
  geom_line(size = 1.2) +
  labs(title = "GDP per Capita", y = "USD", x = "Year") +
  theme_minimal()

```

## GDP Analysis

-   Over the six-decade span, Canada maintains the highest GDP per capita, and Canada's consistent upward trajectory reflects stable economic growth, while Cuba exhibits both lower levels and greater variability.
-   New Zealand closely follows but at a lower level.
-   Cuba’s GDP per capita is significantly lower and more volatile.

This reflects differences in political systems and access to international markets.

## Life Expectancy at Birth

Life expectancy is a core measure of population health.\
The following table compares average life expectancy across the three countries.

```{r}
life <- all_data[Indicator == "Life expectancy at birth, total (years)"]

# Summary statistics by country
life[, .(AvgLife = mean(Value, na.rm=TRUE)), keyby = Country]

```

## Plot (Life Expectancy at Birth):

This plot shows the evolution of life expectancy from 1960 to recent years.

```{r}
ggplot(life, aes(x = Year, y = Value, color = Country)) +
  geom_line(size = 1.2) +
  labs(title = "Life Expectancy at Birth", y = "Years", x = "Year") +
  theme_light()

```

## Fertilizer Consumption

This indicator reveals how intensively countries use fertilizers in agriculture.\
Cuba shows extremely high usage, which may reflect policy differences or data irregularities.

```{r}
fert <- all_data[Indicator == "Fertilizer consumption (% of fertilizer production)"]

fert[, .(MeanValue = mean(Value, na.rm=TRUE)), keyby = .(Country)]

```

## Plot (Fertilizer Consumption):

```{r}
ggplot(fert, aes(x = Year, y = Value, color = Country)) +
  geom_line() +
  labs(title = "Fertilizer Consumption", y = "% of production", x = "Year")

```

Cuba's fertilizer consumption stands out with values 5–10 times higher than the other countries.

## Conclusion

-   GDP and life expectancy trends match global expectations.
-   Cuba shows anomalies in fertilizer consumption—requires further investigation.
-   Future work could expand the scope to include climate-related or education indicators across a broader set of countries.

## Custom Styling:

Font style: applied via individuation.css

Clean theme

Slide numbers and TOC included
